---
title: "demo"
author: "Yifan Zhao"
date: "13/05/2022"
output: md_document
  
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
getwd()
setwd("/Users/yvonne/Downloads/PHP2650SL/php2650_project")
knitr::opts_chunk$set(echo = TRUE)
# html_document:
#     df_print: paged

```


# Veteran Data

In this demo we're using 'veteran' data from survival package, which records data of randomized trial of two treatment regimens for lung cancer. 
```{r}
read.csv('figures/veteran.csv')  %>% kable(caption = 'Veteran Data')
```

First, we split data into training set and test set:
```{r}
library(survival)
library(data.table)
library(ranger)
library(caret)

set.seed(123456)

data(veteran)
vet = veteran
vet$T = Surv(time=vet$time, event=vet$status)

dt <- data.table(veteran)

# Next, we split the data in a training and test set.
ind <- sample(1:nrow(dt),round(nrow(dt) * 0.7,0))
dt_train <- dt[ind,]
dt_test <- dt[!ind,]
vet.tr = vet[ind,]
vet.te = vet[-ind,]
```


Use Kaplan Meier Curve as a visual summary of the survival probability between treatment groups:

```{r}
# plot survival curve:
kmvet = survfit(T~trt, data=vet)
# km curve
plot(kmvet, col=c('blue','red'), xlab='Time', ylab='Survival Probability', main='Kaplan Meier Curves')
legend("topright", lwd = 1, col = c('blue','red'), cex=0.7, y.intersp = 0.5,
       legend = c('trt=1', 'trt=2'))
abline(h=0.5,lty=3)

```


## Cox PH Model

We first fit a COX PH Model. Using backward selection, we found the 'best' variable to fit the data. However, the log-log survival curves show non-parallel curves between variable groups. So the PH assumption might be violated. 

```{r, results=FALSE}
########################################  COX PH ####################################
coxm0 = coxph(T~(celltype+trt+karno+diagtime+age+prior)^2, data=vet.tr, ties='breslow')
# coxm1 = step(coxm0, direction = "backward")

# model selected from backward selection
coxm1 = coxph(formula = T ~ celltype + trt + karno + diagtime + age +
    prior + celltype:diagtime + trt:karno + trt:prior + karno:diagtime +
    karno:age, data = vet.tr, ties = "breslow")
# summary(coxm1)

# PH assumption: non-parallel, violated
par(mfrow=c(2,2))
sapply(list(vet.tr$trt, vet.tr$celltype, vet.tr$karno, vet.tr$prior), 
       function(var) plot(survfit(T ~ var, data=vet.tr),
                          col=1:10,
                          fun="cloglog",
                          ylab='log(-log(S(t)))', 
                          xlab='log(time)',
                          xlim=c(20,250),
                          ylim=c(-2,0.5),
                          main='Log-Log Survival Curves')
)
```

The goodness of fit test based on Schoenfeld Residuals is another way to test the PH assumption, which is the null hypothesis. However, we can see a lot of p values of less than 0.05, this validated our concern that the proportional hazard assumption is violated.

```{r}
cox.zph(coxm1) 
```


## Random Survival Models

### Use ranger package to train the random survival models. 

Recall that in random forest, we need to find number of variables **mtry ** to randomly select from at each node.  First, we find the optimal mtry parameter that gives the smallest OOB error. Over 1000 iterations, m = 3 has the highest frequency of giving the smallest OOB error. 

```{r}
########################################  Ranger #############################################
par(mfrow=c(1,1))

# I created a simple function to find best m:
functune = function(m){
  sapply(1:6, function(m) ranger(Surv(time, status) ~ .,
                data = dt_train,
                mtry = m,
                verbose = TRUE,
                write.forest=TRUE,
                num.trees= 1000,
                importance = 'permutation')$prediction.error)
}
# I commented out this line of code to save running time: 
# findM = apply(replicate(100,functune()), 2, which.min)
# save(findM, file='findM')
load(file='findM')
# m = 3 is the optimal m
# table(findM) 

```



So we fit random survival tree with m = 3. 

```{r}
# fit random survival tree
r_fit <- ranger(Surv(time, status) ~ .,
                data = dt_train,
                mtry = 3,
                verbose = TRUE,
                write.forest=TRUE,
                num.trees= 1000,
                importance = 'permutation')
```


The model gives the fitted survival probability in a table format, where rows represent individual patient and columns represent event time points, from beginning to the end. 

```{r}
# distinct survival time in training data
# r_fit$unique.death.times

# fitted survival: survival probability
# rows represent individual patient; columns represent event time points
fit.surv = r_fit$survival
fit.surv %>% data.frame() %>% head(5)
```

Then I samples 4 individuals to plot fitted survival curve.

```{r}
par(mfrow=c(1,1))
set.seed(12345)
individual = sample(1:dim(dt_train)[1], 4)
plot(r_fit$unique.death.times, fit.surv[individual[1],], type = 'l', col = 'red',xlab='Time', ylab='Survival Probability', main='Fitted Survival Curve (4 individuals)')
lines(r_fit$unique.death.times,fit.surv[individual[2],], type = 'l', col = 'blue',xlab='Time', ylab='Survival Probability')
lines(r_fit$unique.death.times,fit.surv[individual[3],], type = 'l', col = 'green',xlab='Time', ylab='Survival Probability')
lines(r_fit$unique.death.times,fit.surv[individual[4],], type = 'l', col = 'pink',xlab='Time', ylab='Survival Probability')
abline(v=dt_train[individual,]$time, lty=3, col=c('red','blue','green','pink'))
legend("topright", lwd = 1, col = c('red','blue','green','pink'),
       legend = c('id14', 'id51', 'id80', 'id90'))

dt_train[individual,]
```


The variable importance obtained here is very similar to what Cox PH model gave - cell type, treatment, karno, diagtime, age and prior are all considered as important predictor of survival time, except that Cox PH model is able to show interactions between the variables.

```{r}
r_fit$variable.importance
```


In the following chunk of code, I calculated the accuracy of the model when it needs to predict survival after 61 days (this is purely arbitrary for demonstrational purposes).

```{r}
preds <- predict(r_fit, dt_test, type = 'response')$survival
pred.surv <- data.table(preds)
colnames(pred.surv) <- as.character(r_fit$unique.death.times)

prediction <- pred.surv$`61` > 0.5
real <- dt_test$time >= 61
# confusion matrix:
table(prediction, real)
```



## GBSG2

https://rdrr.io/cran/TH.data/man/GBSG2.html

A data frame containing the observations from the GBSG2 study.

This data frame contains the observations of 686 women:

horTh
hormonal therapy, a factor at two levels no and yes.
age
of the patients in years.
menostat
menopausal status, a factor at two levels pre (premenopausal) and post (postmenopausal).
tsize
tumor size (in mm).
tgrade
tumor grade, a ordered factor at levels I < II < III.
pnodes
number of positive nodes.
progrec
progesterone receptor (in fmol).
estrec
estrogen receptor (in fmol).
time
recurrence free survival time (in days).
cens
censoring indicator (0- censored, 1- event).


reference: https://arxiv.org/pdf/1612.08974.pdf
```{r}
########################################  GBSG2:  #############################################
# install.packages('ggRandomForest')
library(ggRandomForests)
data("GBSG2", package = "TH.data")
gbs = GBSG2
# ?GBSG2
head(gbs)


dim(gbs)[1]
set.seed(999)
tidx = sample(1:dim(gbs)[1], floor(dim(gbs)[1]*0.25))
gbs.train = gbs[-tidx,]
gbs.test = gbs[tidx,]
# create survival object 
gg_dta <- gg_survival(interval = "time",
                      censor = "cens",
                      by = "horTh",
                      data = gbs.train,
                      conf.int = 0.95 )

# KM curve
plot(gg_dta)+
  labs(y = "Survival Probability", x = "Observation Time", color = "Hormonal Therapy", fill = "Hormonal Therapy") +
  theme(legend.position = c(0.2,0.2))+
  coord_cartesian(y = c(0, 1.01))

# Cumulative Hazard plot:
plot(gg_dta, type ="cum_haz")+
  labs(y = "Cumulative Hazard", x = "Observation Time(years)", color = "Hormonal Therapy", fill = "Hormonal Therapy") +
  theme(legend.position = c(0.2, 0.8))+
  coord_cartesian( ylim = c(-0.02, 1.22))

# grows the forest
rfsrc.gbs <- rfsrc(Surv(time, cens) ~ ., data = gbs.train,
                     nsplit = 10, na.action = "na.impute",tree.err = TRUE, importance = TRUE)
print.rfsrc(rfsrc.gbs)

# plot OOB error
plot(gg_error(rfsrc.gbs))


# #对测试集的预测
# predict.rfsrc(rfsrc_pbcmy, data = pbc.test)
# predict.rfsrc(rfsrc_pbcmy, data = pbc.trial)
# #VIMP法变量筛选
# plot(gg_vimp(rfsrc_pbcmy))
# #最小深度法变量筛选
# varsel_pbc <- var.select(rfsrc_pbcmy)
# gg_md <- gg_minimal_depth(varsel_pbc)
# plot(gg_md)
# #两种方法相结合的散点图
# plot(gg_minimal_vimp(gg_md))+theme(legend.position = c(0.8,0.2))
# plot(gg_minimal_vimp(rfsrc_pbcmy))+theme(legend.position = c(0.8,0.2))
# gg_minimal_vimp(rfsrc_pbcmy)

```

